Missing values

### 2021-05-25 missing_values.Rmd
update notebooks for quaterly meeting

---
title: "Missing values"
author: "Juan Rocha"
date: "May 2021"
output:
  html_notebook:
      toc: true
      toc_float: true
      highlight: tango
      code_folding: hide
      theme: 
          code_font:
            google: Fira Code
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(naniar)
library(plotly)
library(patchwork)

theme_set(theme_light())
```

@@

```{r}
ecofp %>% 
    ggplot(aes(year, value)) +
    geom_line(aes(color = shortName), show.legend = FALSE) +
    facet_wrap(~record, scales = "free_y" ) +
    theme_light(base_size = 9)
```

@@

```{r}
g1 <- ecofp %>%
    filter(record == "EFConsPerCap") %>%
    select(countryName, year, value) %>% 
    pivot_wider(id_col = countryName, names_from = year, values_from = value) %>% 
    pivot_longer(cols = 2:last_col(), names_to = "year", values_to = "value") %>% 
    mutate(missing = is.na(value), year = as.numeric(year)) %>% 
    group_by(year) %>% 
    summarise(n_nas  = sum(missing)) %>% 
    ggplot(aes(year, n_nas)) + 
    geom_col(aes(fill = (year>1995)), alpha = 0.5) + 
    coord_flip() +
    theme_light(base_size = 7) +
    theme(legend.position = c(0.8,0.8)) + labs(title = "Missing values per year")

g2 <- ecofp %>%
    filter(record == "EFConsPerCap", year > 1995) %>%
    select(countryName, year, value) %>% 
    pivot_wider(id_col = countryName, names_from = year, values_from = value) %>% 
    pivot_longer(cols = 2:last_col(), names_to = "year", values_to = "value") %>% 
    mutate(missing = is.na(value), year = as.numeric(year)) %>% 
    group_by(countryName) %>% 
    summarise(n_nas  = sum(missing)) %>% 
    filter(n_nas > 0) %>% 
    mutate(countryName = as_factor(countryName) %>% fct_reorder(., n_nas)) %>% 
    ggplot(aes(countryName, n_nas)) + 
    geom_col(aes(fill = (n_nas < 11)), alpha = 0.5) + coord_flip() +
    theme_light(base_size = 7) + 
    labs(title = "Missing values per country after 1995") +
    theme(legend.position = c(0.8, 0.1))

@@

```{r}
inq <- readxl::read_xlsx(
    path = "~/Documents/Projects/DATA/WIID_World_Income_Inequality_DB/WIID_06MAY2020.xlsx")

inq %>% 
    #filter(resource == "Income (net/gross)") %>% 
    ggplot(aes(year, gini_reported)) +
    geom_line(aes(color = country), show.legend = FALSE) +
    facet_grid(region_un ~ resource_detailed ) +
    theme_light(base_size = 6)

```
Need to read the docs carefully, it seems many measurements per year and it's hard to tell what is the unit per time that we should use for the analysis.

```{r}
inq %>% 
    ggplot(aes(year, gini_reported)) +
    geom_line(aes(color = country), show.legend = FALSE) +
    facet_grid(region_un ~ resource)
```
It seems that `Income (net)` is the variable with most observations. Let's see:

```{r}
g <- inq %>% 
    filter(year > 1950, resource == "Income (gross)") %>% 
    ggplot(aes(year, gini_reported)) +
    geom_line(aes(color = country), size = 0.1, show.legend = FALSE) +
    facet_wrap(~resource_detailed)

g
```
```{r}
inq %>% 
    filter(year > 1950, resource_detailed == "Income, gross") %>% 
    ggplot(aes(gini_reported, palma)) +
    geom_miss_point(size = 1, alpha = 0.5) +
    theme(legend.position = c(0.1,0.8))
```

@@

```{r fig.width=7}
g <- inq %>% 
    filter(!is.na(gini_reported), year > 1990) %>% select(country, year, gini_reported) %>% 
    group_by(country, year) %>% 
    summarize(gini_mean = mean(gini_reported, na.rm = TRUE)) %>% 
    pivot_wider(id_cols = country, names_from = year, values_from = gini_mean) %>% 
    pivot_longer(2:last_col(), names_to = "year", values_to = "gini_mean") %>%
    mutate(year = as.numeric(year), missing = is.na(gini_mean)) %>% 
    ggplot(aes(year, gini_mean)) +
    geom_line(aes(color = country), show.legend = FALSE, size = 0.5) 
    # geom_miss_point(show.legend = FALSE)
ggplotly(g)
```

@@

>>>>

---
title: "Missing values"
author: "Juan Rocha"
date: "May 2021"
output:
  html_document:
    toc: yes
    toc_float: yes
    highlight: tango
    code_folding: hide
    df_print: paged
    theme: 
      bootswatch: cosmo
      code_font:
        google: Fira Code
    self_contained: true

  html_notebook:
    toc: yes
    toc_float: yes
    highlight: tango
    code_folding: hide
    df_print: paged
    theme: 
      bootswatch: cosmo
      code_font:
        google: Fira Code
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache = TRUE, warning = FALSE, message = FALSE, echo = TRUE,
  fig.width = 8, fig.height = 5
)

library(tidyverse)
library(naniar)
library(plotly)
library(patchwork)

theme_set(theme_light(10))
```

@@

```{r}
ecofp %>% 
    ggplot(aes(year, value)) +
    geom_line(aes(color = shortName), show.legend = FALSE) +
    facet_wrap(~record, scales = "free_y" ) 
```

@@

```{r fig.height=3, fig.width=8}
g1 <- ecofp %>%
    filter(record == "EFConsPerCap") %>%
    select(countryName, year, value) %>% 
    pivot_wider(id_col = countryName, names_from = year, values_from = value) %>% 
    pivot_longer(cols = 2:last_col(), names_to = "year", values_to = "value") %>% 
    mutate(missing = is.na(value), year = as.numeric(year)) %>% 
    group_by(year) %>% 
    summarise(n_nas  = sum(missing)) %>% 
    ggplot(aes(year, n_nas)) + 
    geom_col(aes(fill = (year>1995)), alpha = 0.5) + 
    coord_flip() +
    theme_light(base_size = 7) +
    theme(legend.position = c(0.8,0.8)) + labs(title = "Missing values per year")

g2 <- ecofp %>%
    filter(record == "EFConsPerCap", year > 1995) %>%
    select(countryName, year, value) %>% 
    pivot_wider(id_col = countryName, names_from = year, values_from = value) %>% 
    pivot_longer(cols = 2:last_col(), names_to = "year", values_to = "value") %>% 
    mutate(missing = is.na(value), year = as.numeric(year)) %>% 
    group_by(countryName) %>% 
    summarise(n_nas  = sum(missing)) %>% 
    filter(n_nas > 0) %>% 
    mutate(countryName = as_factor(countryName) %>% fct_reorder(., n_nas)) %>% 
    ggplot(aes(countryName, n_nas)) + 
    geom_col(aes(fill = (n_nas < 11)), alpha = 0.5) + coord_flip() +
    theme_light(base_size = 7) + 
    labs(title = "Missing values per country after 1995") +
    theme(legend.position = c(0.8, 0.2))

@@

```{r}
inq <- readxl::read_xlsx(
    path = "~/Documents/Projects/DATA/WIID_World_Income_Inequality_DB/WIID_06MAY2020.xlsx")
inq
```
Need to read the docs carefully, it seems many measurements per year and it's hard to tell what is the unit per time that we should use for the analysis.

```{r}
inq %>% 
    ggplot(aes(year, gini_reported)) +
    geom_line(aes(color = country), show.legend = FALSE) +
    facet_grid(region_un ~ resource)
```

```{r}
inq %>% 
    filter(year > 1950, resource_detailed == "Income, gross") %>% 
    ggplot(aes(gini_reported, palma)) +
    geom_miss_point(size = 1, alpha = 0.5) +
    theme(legend.position = c(0.1,0.8))
```

@@

```{r fig.width=8}
g <- inq %>% 
    filter(!is.na(gini_reported), year > 1990) %>% select(country, year, gini_reported) %>% 
    group_by(country, year) %>% 
    summarize(gini_mean = mean(gini_reported, na.rm = TRUE)) %>% 
    pivot_wider(id_cols = country, names_from = year, values_from = gini_mean) %>% 
    pivot_longer(2:last_col(), names_to = "year", values_to = "gini_mean") %>%
    mutate(year = as.numeric(year), missing = is.na(gini_mean)) %>% 
    ggplot(aes(year, gini_mean)) +
    geom_line(aes(color = country), show.legend = FALSE, size = 0.5) 
ggplotly(g, tooltip = "country") 
```

@@

## Analysis

The bottle neck is centaintly our proxies of inequality, no matter what we use. Below a mock up example with some arbitrary choices.

```{r}

dat <- ecofp %>%
  filter(record == "EFConsPerCap") %>%
  select(countryName, isoa2, year, value) %>% 
  rename(EFconsPerCap = value, country = countryName) %>% 
  filter(year >= 1990)

dat <- dat %>% 
  left_join(
    .,
    inq %>% 
      filter(!is.na(gini_reported), year >= 1990) %>% 
      select(country, isoa2 = c2, country_code = c3, year, gini_reported) %>% 
      group_by(country, year, isoa2, country_code) %>% 
      summarize(gini_mean = mean(gini_reported, na.rm = TRUE))

  )

dat <- dat %>% 
  left_join(
    ., 
    dat1 %>% 
      select(-x33) %>% 
      filter(indicator_name == "GNI per capita (current US$)") %>% 
      pivot_longer(cols = x1990:x2017, names_to = "year", values_to = "value") %>%
      mutate(year = str_remove(year, "x")) %>% 
      mutate(year = as.numeric(year)) %>% 
      mutate(ismissing = is.na(value)) %>% 
      filter(year >= 1990) %>% 
      rename(country = country_name, gni = value)
  )

dat

```
Each path is a country trajectory over time for the data we have available since 1990. Here are the three phase space graphs, one for each combination of variables in the trilemma. Ideally, we want to understand if being in a particular corner of the space force you to certain trajectories in other dimensions (e.g. being poor country forces you to treat the environment badly while you reach a minimum of equality or economic activity). The red dots are missing values and remind us why it is difficult to perform a full dynamic analysis, and many dots that are not missing are left alone, mining it will be hard to reconstruct time series for some countries.

```{r}
p1 <- dat %>% 
  ggplot(aes(EFconsPerCap, gini_mean, group = country)) +
  geom_miss_point(size = 0.5) +
  geom_path(size = 0.2, color = "gray50") +
  theme(legend.position = c(0.8, 0.8))

p2<- dat %>% 
  ggplot(aes(gni, gini_mean, group = country)) +
  geom_miss_point(size = 0.5, show.legend = FALSE) +
  geom_path(size = 0.2, color = "gray50") +
  scale_x_log10()

p3<- dat %>% 
  ggplot(aes(gni, EFconsPerCap, group = country)) +
  geom_miss_point(size = 0.5, show.legend = FALSE) +
  geom_path(size = 0.2, color = "gray50") +
  scale_x_log10()

p1 + p2 + p3
```

```{r}
# g <- dat %>% 
#   ggplot(aes(EFconsPerCap, gini_mean, group = country)) +
#   geom_point(aes(color = country), size = 0.5, alpha = 0.5) +
#   geom_path(aes(color = country), size = 0.3) +
#   theme_light(base_size = 7)
# 
# ggplotly(g, tooltip = c("country", "gini_mean", "EFconsPerCap"),
#          width = 640, height = 480)
```

```{r}
g <- dat %>% 
  ggplot(aes(gni, gini_mean, group = country)) +
  geom_point(aes(color = country), size = 0.5, alpha = 0.5) +
  geom_path(aes(color = country), size = 0.3) +
  scale_x_log10() +
  theme_light(base_size = 7)

ggplotly(g)
```

```{r}
g <- dat %>% 
  ggplot(aes(gni, EFconsPerCap, group = country)) +
  geom_point(aes(color = country), size = 0.5, alpha = 0.5) +
  geom_path(aes(color = country), size = 0.3) +
  scale_x_log10()+
  theme_light(base_size = 7)

ggplotly(g)
```
